 

### 2021-06-12 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

Exposed-sun-leaves, though equipped with thermoregulatory traits are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] where although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steeply in sun-leaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006]. However, $T_{opt}$ of $gs$ doesnot vary between sun and shade leaves [Slot et al. 2019]. The diurnal $gs$ pattern and resulting $A_{max}$ varies between species, for example, $gs$ is greater in angiosperms than gymnosperms [@bonan_ecological_2016; @kleinStomatalSensitivityCO22019].
### Photosynthesis

@@

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Biophysically, we might expect sun-leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5 in Campbell and Norman 1998; *revisit this!*), where earlier evidence showed that $T_{opt}$ of $J_{max}$ increased with irradiance in an upper-canopy leaf compared to shaded lower-canopy leaf [@niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. However, such a trend is not apparent among recent studies, where for 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al. 2019 ; @hernandezSimilarTemperatureDependence2020]. Similarly, $T_{opt}$ of $V_{max}$ and $J_{max}$ didnot vary for tropical sun- and shade-leaves  [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], and along the vertical gradient of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].

Interestingly, within canopy Japanese oak sun and shade leaves under respective peak light conditions were found to maintain similar rate of excess energy production. Excessive light energy not utilized can cause photoinhibition.Thus, sun leaves utilize this excess energy through high electron transport rate ($ETR$) and thermal dissapation, whereas shade leaves compensate for low $ETR$ by adjusting their $N_{area}$ to maintain high photosythetic capacity than those required to maximize C net gain, a suggested mechanism for cirucmventing photoinihition and thermal dissaptation under sunflecks [@kitaoLeavesJapaneseOak2012; @kitaoCanopyNitrogenDistribution2018]. Perhaps, differences in these thermal stablization strategies realizes similar $T_{opt}$ for photosynthesis in sun and shade leaves.

@@

A few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with significant differences among species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], whereas Griffin et al. 2002 found that in Populas deltoides, lower canopy leaf respiration was more sensitive to changes in $T_{air}$ than upper canopy. Other studies have observed no difference in $Q_{10}$ across species vertical profiles and seedling sun vs. shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008], although at Coweeta, temperate trees showed a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased [@bolstad_foliar_1999](*see their Fig. 1*). (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so. May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*). @bolstad_foliar_1999 also found evidence of acclimation, but there remained a declining trend of respiration with elevation.
Furthermore, response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] showed no variation in canopy positions of Quercus rubra, whereas on considering site differences, canopy-position effect in respiration was much greater in mesic sites, and seasonal effect was greater in drier sites [@xuSeasonalVariationTemperature2006].

@@

Canopies are crucial to microclimatic functioning, as they buffer several aspects of the microenvironment [Zelleweger et al. 2019]. This buffering is influenced by the hydraulogical-cycle at large [Davis et al. 2019], therefore severe droughts place canopies at risk, particularly larger trees that suffer more during droughts[Bennett et al. 2015]. Further, increase in global temperatures has shown that canopies of temperate and tropical forests function close to thermal threshold and beyond photosynthetic optimas [@mau_temperate_2018; @huangAirTemperatureOptima2019]. Increase in such canopy $T_{leaf}$ reduces forest GPP [Pau et al. 2018], specifically in the tropics where thermal limits are narrower than temperate forests [Perez and Feeley, 2018]. Further, warm-season associated with increased VPD, along with drought exacerbates tree stress at a regional scale, with implications for forest structure [@parkwilliamsTemperaturePotentDriver2013]
[**to include Teskey 2015; Breashers et al. 2021**]

@@

While climate change driven forest shifts are observed, certain forests are more resilient to warming and droughts than others, such as African rainforests [Amy Bennett's new paper]. Forest resiliency is also observed through increasing temperature acclimation among tropical and boreal species [martijn's papers; Danielle Way's papers; Atkin's papers] and reversible recovery of canopy heat stress [Smith et al. 2020] among other mechanisms [for eg. *REF*]

>>>>

Exposed-sun-leaves, though equipped with thermoregulatory traits are largely constrained by $gs$ than shade-leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020] where although evapotranspiration and $gs$ peak during the day, during the midday and afternoon $gs$ declines more steeply in sun-leaves resulting in increased $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004; @kochDiurnalPatternsLeaf1994b; @kosugiSeasonalFluctuationsTemperature2006]. However, $T_{opt}$ of $gs$ doesnot vary between sun and shade leaves [Slot et al. 2019]. The diurnal $gs$ pattern and resulting $A_{max}$ varies between species, for example, $gs$ is greater in angiosperms than gymnosperms [@bonan_ecological_2016; @kleinStomatalSensitivityCO22019].

### Photosynthesis

@@

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Biophysically, we might expect sun-leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5 in Campbell and Norman 1998; *revisit this!*; @niinemetsShapeLeafPhotosynthetic1999; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. However, such a trend is not apparent among recent studies, where for 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al. 2019 ; @hernandezSimilarTemperatureDependence2020]. Similarly, $T_{opt}$ of $V_{max}$ and $J_{max}$ didnot vary for tropical sun- and shade-leaves  [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], and along the vertical gradient of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].

Interestingly, within canopy Japanese oak sun and shade leaves were found to maintain similar rate of excess energy production under respective peak light conditions. Excessive light energy not utilized can cause photoinhibition. Thus, sun leaves utilize this excess energy through high electron transport rate ($ETR$) and thermal dissapation, whereas shade leaves compensate for low $ETR$ by adjusting their $N_{area}$ to maintain high photosythetic capacity than those required to maximize C net gain, a suggested mechanism for cirucmventing photoinihition and thermal dissaptation under sunflecks [@kitaoLeavesJapaneseOak2012; @kitaoCanopyNitrogenDistribution2018]. Perhaps, differences in these thermal stablization strategies realizes similar $T_{opt}$ for photosynthesis in sun and shade leaves.

@@

A few studies have found $Q_{10}$ to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with significant differences among species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996], whereas Griffin et al. 2002 found that in Populas deltoides, lower canopy leaf respiration was more sensitive to changes in $T_{air}$ than upper canopy. Other studies have observed no difference in $Q_{10}$ across species vertical profiles and seedling sun vs. shade leaves [@weerasingheCanopyPositionAffects2014; @bolstad_foliar_1999; @zaragoza-castellsDoesGrowthIrradiance2007; @zaragoza-castellsClimatedependentVariationsLeaf2008], although at Coweeta, temperate trees showed a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased [@bolstad_foliar_1999](*see their Fig. 1*). (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so. May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*). @bolstad_foliar_1999 also found evidence of acclimation, but there remained a declining trend of respiration with elevation.
Furthermore, response of respiration to $T_{leaf}$ [partially represented  by $E_{0}$] also showed no variation in canopy positions of Quercus rubra, whereas on considering site differences, canopy-position effect in respiration was much greater in mesic sites, and seasonal effect was greater in drier sites [@xuSeasonalVariationTemperature2006].

@@

Canopies are crucial to microclimatic functioning, as they buffer several aspects of the microenvironment [Zelleweger et al. 2019]. This buffering is influenced by the hydraulogical-cycle at large [Davis et al. 2019], therefore severe droughts place canopies at risk, particularly larger trees that suffer more during droughts[Bennett et al. 2015]. Further, increase in global temperatures has shown that canopies of temperate and tropical forests function close to thermal threshold and beyond photosynthetic optimas [@mau_temperate_2018; @huangAirTemperatureOptima2019]. Increase in such canopy $T_{leaf}$ reduces forest GPP [Pau et al. 2018], specifically in the tropics where thermal limits are narrower than temperate forests [Perez and Feeley, 2018]. A combination of warm-season associated with increased VPD, along with drought exacerbates tree stress at a regional scale, with implications for forest structure [@parkwilliamsTemperaturePotentDriver2013]
[**to include Teskey 2015; Breashers et al. 2021**]

@@

In the face of global climate change, certain forests, such as African rainforests [Amy Bennett's new paper] are more resilient to warming and drought than other temperate and tropical forests. Additionally, increasing temperature acclimation among tropical and boreal species [martijn's papers; Danielle Way's papers; Atkin's papers] and reversible recovery of canopy heat stress [Smith et al. 2020] among other mechanisms [for eg. *REF*], point towards rays of hope.
